 

### 2021-07-11 vertical-thermal-review.Rmd
remove not on @SlotMartijn 's data
I think it's late to try to include this, but if @SlotMartijn feels otherwise we can change course here. For now, I'm taking it out and will close the issue (#8)
Update vertical-thermal-review.Rmd

Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by other biophysical drivers (reviewed above) and by leaf traits and stomatal conductance (see following sections), with leaves adjusted to their environment so as to maintain, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. 
Leaves can be either cooler than air at high $T_{air}$ or warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ [megathermy, @doughtyAreTropicalForests2008].
While $T_{leaf}$ is most commonly within a few degrees of $T_{air}$ [poikilothermy, @cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020], it is rarely exactly equal.

@@

Further, higher heat tolerance in exposed canopies is facilitated by isoprenes in species capable of emitting isoprenes, *providing these species with a selective advantage in a warming world* [*REF*].

@@

Leaf age affects vulnerability to heat stress independent of light environment [e.g., Marias et al. 2017], and has a strong effect on metabolism [@kikuzawaSynthesisRelationshipsLeaf2006], including photosynthesis and respiration, and their temperature sensitivities [@zhouLeafageEffectsTemperature2015; @niinemetsLeafAgeDependent2016].
Sun leaves have shorter lifespans than shade leaves, and thus upper canopy leaves are on average younger than subcanopy leaves, therefore they operate closer to their photosynthetic optimum. However, because leaf turnover rates of sun leaves in the upper canopy are faster than in the subcanopy, less damage  accumulates [@millerOnlySunlitLeaves2021].
Once leaves leaves are fully expanded and mature, heat tolerance decreases with leaf age [@zhouLeafageEffectsTemperature2015; REFS], as do $g_s$ and photosynthesis [with a few exceptions between species, @niinemetsLeafAgeDependent2016]. 
In evergreen species, temperature responses of photosynthesis and temperature sensitivity of respiration ($Q_{10}$) have also been observed to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].
Additionally, because older leaves reach maximum $T_{leaf}$ sooner than younger leaves and can occupy a large part of the canopy, this would affect vertical gradient of thermal sensitivity [@zhouLeafageEffectsTemperature2015] **this needs some confirming by co-authors**.

@@

Generally, the pattern of fundamental trait adaptation to light gradients across forest species is similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. 
However, the range in trait values can be several times greater in multi-species canopy than in understory species, based on their inherent species-specific traits and responses to environmental conditions. 
Spatio-temporal gradients offer another axis for trait variation, for instance in early- and late-successional species [Ishii and Asano, 2007, @ishiiExploringRelationshipsCanopy2004a]. 
Therefore, $T_{leaf}$ patterns witnin forest ecosystems are underpinned by both intra- and inter-specific variation in traits. 
*For example, species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase than species with larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020].*

##### (*Deciduousness* -- drop header, probably integrate this into paragraph above, or somehow re-organize the whole itnerspecific variation section)

@@

@@

A few studies have found $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], whereas @griffinCanopyPositionAffects2002 found that in *Populas deltoides*, lower canopy leaf respiration was more sensitive to changes in $T_{air}$ than upper canopy. Other studies have observed no difference in $Q_{10}$ within species vertical profiles and in seedling sun vs. shade leaves [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008]. (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so. May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*). 
@xuSeasonalVariationTemperature2006 found that the canopy-position effect gradient in respiration rates was much greater for Quercus rubra trees in mesic than in xeric sites, but $Q_{10}$ did not differ between canopy positions.

@@

>>>>

Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by other biophysical drivers (reviewed above) and by leaf traits and stomatal conductance (see following sections), with leaves adjusted to their environment so as to maintain, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. 
Leaves can be either cooler than air at high $T_{air}$ or warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ [megathermy, @doughtyAreTropicalForests2008].
While $T_{leaf}$ is most commonly within a few degrees of $T_{air}$, it is rarely exactly equal [poikilothermy, @cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].

@@

Further, higher heat tolerance in exposed canopies is facilitated by isoprenes in species capable of emitting isoprenes [REF].

@@

Leaf age affects vulnerability to heat stress independent of light environment [e.g., Marias et al. 2017], and has a strong effect on metabolism [@kikuzawaSynthesisRelationshipsLeaf2006], including photosynthesis and respiration, and their temperature sensitivities [@zhouLeafageEffectsTemperature2015; @niinemetsLeafAgeDependent2016].
Sun leaves have shorter lifespans than shade leaves, and thus upper canopy leaves are on average younger than subcanopy leaves, therefore they operate closer to their photosynthetic optimum. However, because leaf turnover rates of sun leaves in the upper canopy are faster than in the subcanopy, less damage  accumulates [@millerOnlySunlitLeaves2021].
Once leaves leaves are fully expanded and mature, heat tolerance decreases with leaf age [@zhouLeafageEffectsTemperature2015; REFS], as do $g_s$ and photosynthesis [with a few exceptions between species, @niinemetsLeafAgeDependent2016]. 
Additionally, because older leaves reach maximum $T_{leaf}$ sooner than younger leaves and can occupy a large part of the canopy, this would affect vertical gradient of thermal sensitivity [@zhouLeafageEffectsTemperature2015] **this needs some confirming by co-authors**.

@@

Generally, the pattern of fundamental trait adaptation to light gradients across forest species is similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. 
However, the range in trait values can be several times greater in multi-species canopy than in understory species, based on their inherent species-specific traits and responses to environmental conditions. Spatio-temporal gradients offer another axis for trait variation, for instance in early- and late-successional species [Ishii and Asano, 2007, @ishiiExploringRelationshipsCanopy2004a].

@@

Therefore, $T_{leaf}$ patterns within forest ecosystems are underpinned by both intra- and inter-specific variation in traits.

@@

A few studies have found $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], whereas @griffinCanopyPositionAffects2002 found that in *Populas deltoides*, lower canopy leaf respiration was more sensitive to changes in $T_{air}$ than upper canopy. Other studies have observed no difference in $Q_{10}$ within species vertical profiles and in seedling sun vs. shade leaves [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008]. 
@xuSeasonalVariationTemperature2006 found that the canopy-position effect gradient in respiration rates was much greater for Quercus rubra trees in mesic than in xeric sites, but $Q_{10}$ did not differ between canopy positions.

-In evergreen species, temperature responses of photosynthesis and temperature sensitivity of respiration ($Q_{10}$) have also been observed to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].

@@

*For example, species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase than species with larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020].*

-Isoprenes *providing emitting species with a selective advantage in a warming world* [*REF*].
